Dark Control:\\ A Unified Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-04-19 paper/paper.tex
interm

\section{Introduction}
%
When left unperturbed, the human brain is not at rest.
Especially the so-called "default mode network" (DMN) continues
to metabolize large quantities of
oxygen and glucose energy to maintain
neuronal computation
in the absence of behavioral focus
\citep{kenet2003spontaneously, fiser2004small}.
The baseline energy demand is surprisingly weakly modulated
with task demand \citep{raichle2001pnas}. At the opposite, during sleep, the
decoupling of brain structures discarded the idea of the DMN being
only a passive network resonance and rather supported its
important role in sustaining conscious awareness \citep{horovitz2009decoupling}.
This dark matter of brain physiology begs the question
of the biological purpose underlying the DMN
\citep{raichle2006brain}.
What has early been described as the "stream of consciousness"
in psychology \citep{james1890principles}
found a potential neurobiological manifestation
in the DMN
\citep{shul1997, raichle2001pnas}.
This set of some of the most advanced regions in
the human association cortex \citep{mesulam1998sensation, margulies2016situating}
consistently
increase in neural activity during
unfocused everyday mind wandering \citep{raichle2001pnas}.
The DMN may subserve important functions in
higher-order control of human behavior
following the notion of "a hierarchy of brain systems with
the DMN at the top and the salience and dorsal attention systems
at intermediate levels, above thalamic and unimodal sensory
cortex" \citep{carhart2010default}.
In the beginning of the 21st century,
brain imaging may have been the first technique
to allow for the discovery of a unique brain network that subserves
baseline mental activities
\citep{raichle2001pnas, bzdok2015resting}.

\subsection{Towards a formal account of default mode function: higher-order control of the organism}
The human DMN is responsible for a large part of the baseline neural activity,
which typically decreases during psychological experiments.
The standard mode of
neural information maintenance and manipulation has
been argued to mediate evolutionarily conserved functions
\citep{brown1914nature, binder1999conceptual, buzsaki2006rhythms}.
Today, many investigators believe that the DMN implements some form of
probabilistic estimation of past, hypothetical, and
future events
\citep{fox2005, hassabis2007patients, schacter2007remembering, binder2009,
randy2008}.
This brain network
might have emerged to continuously predict environmental events using
mental imagery as an evolutionary advantage.
%
However, information processing in the DMN has also repeatedly
been shown to directly impact human behavior. Goal-directed task performance
improved with decreased activity in default mode areas \citep{weiss2006}
and increased DMN activity was linked to more task-independent,
yet sometimes useful thoughts
\citep{mason2007, seli2016mind}.
%
Gaining insight into the function of the DMN is
particularly challenging because
this network appears to
simultaneously control perception-action cycles and
maintain baseline contemplations
across time, space, and content domains.

The present work takes the perspective of
a human agent faced with the choice of the next actions
and being guided by outcome estimations
of really-happened, hypothetically conceived, and
expected futures to optimize behavioral performance.
A particularly attractive computational framework
to describe, quantify, and predict autonomously acting systems like the brain
is proposed to be the combination of control theory and
reinforcement learning (RL).
It is known that, the more the external world is predictable,
the more mental activity becomes detached from the actual sensory environment
\citep{antrobus1966studies, pope1978regulation}.
Conversely,
the more the ongoing executed task is unknown and unpracticed,
the less stimulus-independent thoughts occur
\citep{filler1973daydreaming, teasdale1995stimulus, christoff2016mind}.
Without requiring explicit awareness,
these ``offline'' processes may however contribute
to optimizing control of the organism.
We formalize
a policy matrix to capture the space of possible actions that can be performed
on the environment given the current state. A value function
maps expected rewards to environmental objects and events.
Switching between states reduces to a sequential processing model.
Informed by outcomes of performed action,
the DMN dynamics are constantly adapted in feedback loops
shaped by prediction error.
A DMN account incorporating RL can naturally embed human behavior
into the tension between exploitative action with immediate gains and
exploratory action with longer-term reward outcomes.

We argue that DMN implication in a
diversity of humans' most advanced cognitive processes
can be parsimoniously recast as prediction error minimization
based on pervasive probabilistic simulations,
thus maximizing action outcome across multiple time scales.
Such a purposeful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of Markov Chain Monte Carlo (MCMC) sampling
\citep{tenenbaum2011grow}.
In this context,
\textit{control} refers to the influence that an agent exerts when interacting
with the environment to encourage preferred states. \textbf{(XXX Guillaume: DM was asking to move this sentence up. Indeed, we may have to define early on the key concepts.}
Even (necessarily imperfect) memory recall,
random day-time mind-wandering, and dreams during sleep
may provide meaningful building blocks to iteratively improve
the predictive DMN machinery to optimize the behavior of the organism.
%
It has indeed been proposed earlier that
the human brain's energy budget is largely dedicated to
``the development and maintenance of [a]
probabilistic model of anticipated events''
\citep{raichle2005intrinsic}.
This idea has been invigorated by
empirical evidence from
neuroscientific experiments \citep{kording2004bayesian, fiser2004small}.
The present paper proposes a
process model that satisfies this contention.

@@

Comparing to the dmPFC,
the vmPFC probably subserves
subjective value evaluation and
risk estimation of relevant environmental stimuli.
This DMN node is more closely associated with
orchestrating adaptive behavior by bottom-up-driven
processing of “what matters now”,
probably drawing on sophisticated value representations
\citep{doherty2015structure}.
Quantitative lesion findings across 344 human individuals confirmed
a substantial impairment in value-based decision making
\citep{glascher2012lesion}.
Indeed,
the vmPFC is preferentially connected with reward-related and limbic areas.
The vmPFC is well known to have monosynaptical connections
with the NAc
in axonal tracing studies in monkeys \citep{haber1995orbital}.
Congruently, the gray-matter volume of the vmPFC and NAc
correlated with indices of value-guided behavior and reward attitudes
in humans
\citep{lebreton2009automatic}.
NAc activity is thought to reflect reward prediction signals from dopaminergic pathways
\citep{schultz1998predictive}
that not only channel action towards basic survival needs but also
subserve RL in humans more broadly \citep{doherty2015structure}.
This is consistent with diffusion MRI tractography in humans and monkeys
\citep{croxson2005quantitative} that
quantified the NAc to
be more connected to the vmPFC than dmPFC in both species.
Two functional connectivity analyses in humans strongly connected
the vmPFC with the NAc, hippocampus (HC),
and PMC.
%
In line with these connectivity findings in animals and humans,
the vmPFC is often proposed to be involved in (external) emotional
reactions and own (visceral) arousal \citep{damasio1996somatic}.
Real or imagined bodily states could be mapped in the vmPFC
as a bioregulatory disposition governing cognition
and decision making.
In neuroeconomic studies of human decision making,
the vmPFC consistently reflects an individual’s subjective
valuation
\citep{behrens2008associative}.
This may be why performance within and across participants
was related to state encoding in the vmPFC.
Such a ``cognitive map'' of the action space was argued to encode
the current task state even when states are unobservable from sensory input,
which was shown to be critical for behavior \citep{Schuck2016}.

@@

RL is a problem-solving technique in which, through interactions with an unknown \textit{environment / world},
an agent learns to reach goals, optimize reward signals, and minimize costs
in an iterative trial-by-error fashion. See Fig. \ref{fig:rl}.
At a given moment, each taken \textit{action} $a$ triggers an \textit{partially observed} change
in the \textit{state} of the environment
$s \rightarrow s'$, accompanied by environmental feedback signals as \textit{reward}
$r = r(s, a)$ (or \textit{punishment},
since it can be negative) collected by the
agent. This is illustrated in Fig. \ref{fig:rl}.
In this view, the environment is partially controlled by
the action of the agent and the reward can be thought
of as satisfaction -- or punishment -- accompanying the execution of
an action.

\begin{wrapfigure}{r}{.4\linewidth}
  \centering
  \includegraphics{rl.png}
  \caption{RL in a nutshell}
  \label{fig:rl}
\end{wrapfigure}

@@

Model-free RL is plausibly realizable in the human brain. Indeed,
it has long been proposed \citep{dayan2008decision} that there
is a rather direct mapping of model-free RL learning algorithms
onto aspects of the brain.
The neurotransmitter dopamine could serve
as a ``teaching signal'' to better estimate value associations
and action policies by controlling
synaptic plasticity at targets such as the ventral and
dorsolateral striatum.
In contrast, \textit{model-based RL} would start off with some mechanistic assumptions about the dynamics of the world.
These can be assumptions about the physical laws governing the agent's world or constraints on the state space and transitions between states.
For instance, if a rat in a maze knows
that standing still will produce no change in the environment, and in particular will not eventually lead to finding the food.
% A core property of human intelligence has been proposed underly to improve
% on an expected utility outcomes as a strategy for action choice in uncertain
% environments \citep{gershman2015computational}.
In our present model-free RL framework,
a rat might represent such knowledge about the world as follows:
\begin{itemize}
\item $r(s, \text{``stand still''}) = 0$ if $s$ does not correspond to a cell / chamber containing food.
\item $p(s'|s,\text{``stand still''}) = 1$ if $s'=s$ and $0$ otherwise.
\item etc.
\end{itemize}

@@

\subsection{Expected values, Bellman equation, and optimal control}
Given only the limited information available from a POMDP, at a state $s$ the average quality or usefulness of taking an action $a$ under a policy $\pi$ is completely captured by the single number
% We can now formulate the purpose of the DMN
% as finding a policy $\pi$ for the human agent that maximizes the
% expected cumulative value of an state-action pair $(a,s)$, also known as the \textit{Q-value}, namely
\begin{equation}
  \label{eq:qval}
  Q^{\pi}(s,a) = \mathbb E [G^\pi|s_0=s,a_0=a],
\end{equation}
called the $Q$-value for the state-action pair $(s,a)$.
In other words, $Q^{\pi}(s,a)$ corresponds to the expected reward
over all possible action trajectories, in which
the agent sets out in the environment in state
$s$, chooses action $a$, and then follows the policy $\pi$ to select future actions.
%% Given that each $\pi$ correspondence to an exhaustive description of
%% behavioral patterns, $Q^{\pi}(s_t,a_t)$ corresponds to the epistemic value of
%% the set of actions given states in finite time.
For the brain,
$Q^{\pi}(s, a)$ defined in \eqref{eq:qval} provides the subjective
utility of executing a specific action; it answers the question
"What is the expected utility of taking action $a$ in this situation ?".
$Q^{\pi}(s,a)$ offers a formalization of optimal behavior that
may well reflect processing aspects of the DMN in human agents.

% One can also define the state-value function

% \begin{eqnarray}
%   \V^\pi(s) := \mathbb E_{a \sim \pi}Q^\pi(s,a) = \mathbb E [G^\pi|s_0=s],
% \end{eqnarray}
% which measures the value of each state $s \in \mathcal S$ under the policy $\pi$. This serves
% as a baseline value for any action. Finally, the advantage function is defined by
% \begin{equation}
% A^\pi(s,a) :=  Q^\pi(s,a) - V^\pi(s).
% \end{equation}

@@

% \subsection{Efficient computation of optimal behavior}
% By construction, maximizing the expectation in  \eqref{eq:qval} is computationally intractable in general.
% This is due to the enormous number of possible action
% trajectories that the agent often needs to
% consider simultaneously.
% % \begin{mdframed}
% %   \textbf{N.B.:}
% %   If the Bellman-based computational ideas have a corresponding
% %   neurobiological implementation, this provides a framework of how
% %   optimal behavior could be computable in the human brain.
% % \end{mdframed}
% One popular solution is called \text{Q-learning}. It remedies
% the issue by only considering the sub-family of
% \textit{deterministic policies}, which map each state to the single best action to
% take at the current state. Such policies define deterministic functions from states to actions.

% \subsubsection{The Bellman equation:  characterizing optimal behavior}
% Q-learning (Watkins \& Dayan, 1992) generates a
% \textit{greedy policy}
Optimal behavior corresponds to a strategy $\pi^*$ for choosing actions such that, for every state, the chosen action guarantees the best possible reward on average. This can be formalized by
% \textbf{XXX please add 1-2 sentences why this is interesting in the current
% context. This statement appears a little alone as it stands.}
\begin{equation}
  \pi^*(s) = \argmax_{a \in \mathcal A}{Q}^*(s, a),
  \label{eq:qlearning}
\end{equation}
where
\begin{equation}
  Q^*(s, a) := \max_{\pi}Q^\pi(s,a),
\end{equation}
the optimal Q-value over all policies. The goal of learning is to approach the policy $\pi^*$ as ``close as possible'', that is to \text{solve} the POMDP. Of course \eqref{eq:qlearning} is just a definition and
represents by no means, a candidate schema for solving POMDPs with even moderately-sized action and state spaces: it's intractable!
% where $\tilde{Q}(s, a)$ is the current estimate of the state-action value
% table.
Fortunately, the \textit{Bellman equation} \citep{sutton1998reinforcement} provides a fixed-point relation which defines $Q^*$ implicitly via a sampling procedure, without the intractable problem
over all every policy, viz
\begin{equation}
  Q^* = \text{Bel}(Q^*),
  \label{eq:bellman}
\end{equation}
where the ``Bellman transform'' $\text{Bel}(Q)$ of an arbitrary $Q$-value function $Q: \mathcal S \times \mathcal A \rightarrow \mathbb R$  is defined by setting
\begin{equation}
  \begin{split}
   \text{Bel}(Q)(s,a) &:=
   \mathbb E_{s' \sim p(s'|s,a)} [r(s,a) + \gamma \max_{a' \in \mathcal A}Q(s', a')]\\
   &= r(s,a) + \mathbb E_{s' \sim p(s'|s,a)} [\gamma \max_{a' \in \mathcal A}Q(s', a')]\\
   &= \text{instantaneous reward} + \text{expected reward for acting greedily thereafter}
%    \; \forall (s,a) \in \mathcal S \times \mathcal A
    % \\&=
    % \mathbb E_{s' \sim p(s'|s,a)} [r(s,a) + \gamma Q^*(s', \pi^*(s'))].
    \end{split}
  \end{equation}
The Bellman equation \eqref{eq:bellman} is a ``self-consistency'' equation which provides
a recursive decomposition of optimal behavior by dividing the $Q$-value function into the immediate
reward and the discounted rewards of the upcoming states.
% The optimal Q-value operator $Q^*$
% is a fixed point for this equation.
As a consequence, the complicated dynamic programming optimization problem \eqref{eq:qlearning}
is broken into simpler sub-problems at different time points.
Exploitation of hierarchical structure in actions consideration
has previously been closely related to the medial prefrontal part of the DMN
\citep{koechlin1999role, braver2002role}.
Using the Bellman equation, each state can be associated with a certain value
to guide action towards a better state, thus improving on the agent's current policy.
% It is a fixed-point equation for the deterministic policy $\pi$.
Note that in \eqref{eq:bellman} the sampling is done only over things which
depend on the environment, and so can be learned off-policy by observing state transitions
triggered by another behavioral policy, which can be stochastic.

\begin{mdframed}
  \textbf{ NCs for the Bellman equation:}
  ...
\end{mdframed}

\subsubsection{Value approximation and greedy action consideration, and experience replay}
%As already mentioned in the previous section, Q-learning optimizes over the class of
% deterministic policies of the form \eqref{eq:qlearning}.
At a time step $k-1$, the world is in a state $s \in \mathcal S$, and the agent greedily takes an
action which it expects to be the must valuable on average, namely

\begin{equation}
  \pi^{\text{greedy}}(s) = \argmax_{a \in \mathcal A}\tilde{Q}(s, a|\theta),
  \label{eq:policy}
\end{equation}
by querying an approximate version $a \mapsto \tilde{Q}(a,s|\theta)$ of the $Q$-value function ``frozen'' at state $s$. In the DMN, The dmPFC (BA9) would implement such a greedy lookup over the action space.
The tensor $\theta$ encodes the model
  parameters and would correspond to synaptic weights and / or connection strengths between
  DMN regions. It is a time-varying neuronal program which dictates how to move from world states $s$ to actions $a$ via the greedy policy \eqref{eq:policy}. Learning consists in rewriting this program on the fly, so that the approximating functional $\tilde{Q}(.,.|\theta)$ becomes closer and closer to the optimum $Q^*$.

The action $a$ taken in \eqref{eq:policy} triggers a reward $r$ collected by the agent,
% (measured on their internal currency of worldly values ?),
after which the environment transitions to a new state $s' \in \mathcal S$.
% \textbf{XXX Please try to smoothen the previous sentence}
At the end of this cycle, a new \textit{experience} $e = (s,a,r,s')$ is produced, which represents an exemplar
behavior of the agent and is recorded in replay memory buffer
(possibly discarding the oldest entries to make space):
$\mathcal D \leftarrow \text{append}(\mathcal D, e)$.
At time step $k$, the agent seeks an update $\theta_{k} \leftarrow \theta_{k-1} + \delta \theta_{k}$ of the parameters for its approximate model of $Q$-value functional. This warrants learning,
and a loss function is in demand. The Bellman equation \eqref{eq:bellman}
provides a way to produce such a loss function \eqref{eq:oracle} as we now outline below.

% A practically successful RL approach
% needs ways to efficiently approximate the state-action value function
% to allow for scaling to large state and action spaces.
Replay consists in sampling (uniform or importance-weighted
\footnote{e.g weighted by TD error of the state transition $s \overset{a}{\rightarrow} s'$.} ?)  mini-batches of experiences
$(s, a, r, s') \sim \mathcal D$ from the replay memory $\mathcal D$.
The agent then tries to approximate
the would-be $Q$-value for the state-action pair $(s,a)$ as predicted by the Bellman equation \eqref{eq:bellman}, namely
\begin{equation}
  y_k := y_k(s,a,s') =  r + \gamma \max_{a'} \tilde{Q}(s', a'|\theta_{k-1}),
\end{equation}
with the prediction of a parametrized regression model $(s,a)
\mapsto \tilde{Q}(s, a|\theta_{k-1})$.
From a neurobiological perspective,
experience replay would be manifested as the re-occurrence of a
neuron spiking sequence that also occurs during specific actions.
%, but the replay has a much faster time scale.
In the human brain, the hippocampus is likely to contribute to such a mechanism,
as neuroscience experiments on various animals have repeatedly indicated
in rats, mice, cats, rabbits, songbirds, and
monkeys \cite{buhry2011,nokia2010,dave2000,skaggs2007}.

>>>>

\section{Introduction}
%
When left unperturbed, the human brain is not at rest.
In the beginning of the 21st century,
brain imaging may have been the first technique
to allow for the discovery of a unique brain network that subserves
baseline mental activities
\citep{raichle2001pnas, bzdok2015resting}.
Especially the so-called "default mode network" (DMN) continues
to metabolize large quantities of
oxygen and glucose energy to maintain
neuronal computation
in the absence of behavioral focus
\citep{kenet2003spontaneously, fiser2004small}.
The baseline energy demand is surprisingly weakly modulated
with task demand \citep{raichle2001pnas}. At the opposite, during sleep, the
decoupling of brain structures discarded the idea of the DMN being
only a passive network resonance and rather supported its
important role in sustaining conscious awareness \citep{horovitz2009decoupling}.
This dark matter of brain physiology begs the question
of the biological purpose underlying the DMN
\citep{raichle2006brain}.
What has early been described as the "stream of consciousness"
in psychology \citep{james1890principles}
found a potential neurobiological manifestation
in the DMN
\citep{shul1997, raichle2001pnas}.
This set of some of the most advanced regions in
the human association cortex \citep{mesulam1998sensation, margulies2016situating}
consistently
increase in neural activity during
unfocused everyday mind wandering \citep{raichle2001pnas}.
The DMN may subserve important functions in
higher-order control of human behavior
following the notion of "a hierarchy of brain systems with
the DMN at the top and the salience and dorsal attention systems
at intermediate levels, above thalamic and unimodal sensory
cortex" \citep{carhart2010default}.

\subsection{Towards a formal account of default mode function: higher-order control of the organism}
The human DMN is responsible for a large part of the baseline neural activity,
which typically decreases during psychological experiments.
The standard mode of
neural information maintenance and manipulation has
been argued to mediate evolutionarily conserved functions
\citep{brown1914nature, binder1999conceptual, buzsaki2006rhythms}.
Today, many investigators believe that the DMN implements some form of
probabilistic estimation of past, hypothetical, and
future events
\citep{fox2005, hassabis2007patients, schacter2007remembering, binder2009,
randy2008}.
This brain network
might have emerged to continuously predict environmental events using
mental imagery as an evolutionary advantage.
%
However, information processing in the DMN has also repeatedly
been shown to directly impact human behavior. Goal-directed task performance
improved with decreased activity in default mode regions \citep{weiss2006}
and increased DMN activity was linked to more task-independent,
yet sometimes useful thoughts
\citep{mason2007, seli2016mind}.
%
Gaining insight into the function of the DMN is
particularly challenging because
this network appears to
simultaneously control perception-action cycles and
maintain baseline contemplations
across time, space, and content domains.

The present work takes the perspective of
a human agent faced with the choice of the next actions
and being guided by outcome estimations
of really-happened, hypothetically conceived, and
expected futures to optimize behavioral performance.
A particularly attractive computational framework
to describe, quantify, and predict intelligent systems, such as the brain,
is proposed to be the combination of control theory and
reinforcement learning (RL).
It is known that, the more the external world is predictable,
the more mental activity becomes detached from the actual sensory environment
\citep{antrobus1966studies, pope1978regulation}.
Conversely,
the more the ongoing executed task is unknown and unpracticed,
the less stimulus-independent thoughts occur
\citep{filler1973daydreaming, teasdale1995stimulus, christoff2016mind}.
Without requiring explicit awareness,
these ``offline'' processes may contribute
to optimizing control of the organism.
We formalize
a policy matrix to capture the space of possible actions that can be performed
on the environment given the current state. A value function
maps expected rewards to environmental objects and events.
Switching between states reduces to a sequential processing model.
Informed by outcomes of performed action,
the DMN dynamics are constantly adapted in feedback loops
shaped by prediction error.
A DMN account incorporating RL can naturally embed human behavior
into the tension between exploitative action with immediate gains and
exploratory action with longer-term reward outcomes.

We argue that DMN implication in a
diversity of humans' most advanced cognitive processes
can be parsimoniously recast as prediction error minimization
based on pervasive probabilistic simulations,
thus maximizing action outcome across multiple time scales.
Such a purposeful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of Markov Chain Monte Carlo (MCMC) sampling
\citep{tenenbaum2011grow}.
In this context,
\textit{control} refers to the influence that an agent exerts when interacting
with the environment to encourage preferred states. \textbf{(XXX Guillaume: DM was asking to move this sentence up. Indeed, we may have to define early on the key concepts.}
Even (necessarily imperfect) memory recall,
random day-time mind-wandering, and dreams during sleep
may provide meaningful building blocks to iteratively improve
the predictive DMN machinery to optimize the behavior of the organism.
%
It has indeed been proposed earlier that
the human brain's energy budget is largely dedicated to
``the development and maintenance of [a]
probabilistic model of anticipated events''
\citep{raichle2005intrinsic}.
This idea has been invigorated by
empirical evidence from
neuroscience experiments \citep{kording2004bayesian, fiser2004small}.
The present paper proposes a
process model that satisfies this contention.

@@

Comparing to the dmPFC,
the vmPFC probably subserves
subjective value evaluation and
risk estimation of relevant environmental stimuli.
This DMN node is more closely associated with
orchestrating adaptive behavior by bottom-up-driven
processing of “what matters now”,
probably drawing on sophisticated value representations
\citep{doherty2015structure}.
Quantitative lesion findings across 344 human individuals confirmed
a substantial impairment in value-based decision making
\citep{glascher2012lesion}.
Indeed,
the vmPFC is preferentially connected with reward-related and limbic regions.
The vmPFC is well known to have monosynaptical connections
with the NAc
in axonal tracing studies in monkeys \citep{haber1995orbital}.
Congruently, the gray-matter volume of the vmPFC and NAc
correlated with indices of value-guided behavior and reward attitudes
in humans
\citep{lebreton2009automatic}.
NAc activity is thought to reflect reward prediction signals from dopaminergic pathways
\citep{schultz1998predictive}
that not only channel action towards basic survival needs but also
subserve RL in humans more broadly \citep{doherty2015structure}.
This is consistent with diffusion MRI tractography in humans and monkeys
\citep{croxson2005quantitative} that
quantified the NAc to
be more connected to the vmPFC than dmPFC in both species.
Two functional connectivity analyses in humans strongly connected
the vmPFC with the NAc, hippocampus (HC),
and PMC.
%
In line with these connectivity findings in animals and humans,
the vmPFC is often proposed to be involved in (external) emotional
reactions and own (visceral) arousal \citep{damasio1996somatic}.
Real or imagined bodily states could be mapped in the vmPFC
as a bioregulatory disposition governing cognition
and decision making.
In neuroeconomic studies of human decision making,
the vmPFC consistently reflects an individual’s subjective
valuation
\citep{behrens2008associative}.
This may be why performance within and across participants
was related to state encoding in the vmPFC.
Such a ``cognitive map'' of the action space was argued to encode
the current task state even when states are unobservable from sensory input,
which was shown to be critical for behavior \citep{Schuck2016}.

@@

RL is a problem-solving technique in which, through interactions with an environment,
an agent learns to reach goals, optimize reward signals, and minimize costs
in an iterative trial-by-error fashion.
At a given moment, each taken \textit{action} $a$ triggers an \textit{partially observed} change
in the \textit{state} of the environment
$s \rightarrow s'$, accompanied by environmental feedback signals as \textit{reward}
$r = r(s, a)$ (or \textit{punishment},
since it can be negative) collected by the
agent. This is illustrated in Fig. \ref{fig:rl}.
In this view, the environment is partially controlled by
the action of the agent and the reward can be thought
of as satisfaction -- or punishment -- accompanying the execution of
an action.

\begin{wrapfigure}{r}{.4\linewidth}
  \centering
  \includegraphics{aeloop-300x183.png}
  \caption{RL in a nutshell}
  \label{fig:rl}
\end{wrapfigure}

@@

Model-free RL is plausible to be realizable in the human brain. Indeed,
it has long been proposed \citep{dayan2008decision} that there
is a rather direct mapping of model-free RL learning algorithms
onto aspects of the brain.
The neurotransmitter dopamine could serve
as a ``teaching signal'' to better estimate value associations
and action policies by controlling
synaptic plasticity at targets such as the ventral and
dorsolateral striatum.
In contrast, \textit{model-based RL} would start off with some mechanistic assumptions about the dynamics of the world.
These can be assumptions about the physical laws governing the agent's world or constraints on the state space and transitions between states.
For instance, if a rat in a maze knows
that standing still will produce no change in the environment, and in particular will not eventually lead to finding the food.
% A core property of human intelligence has been proposed underly to improve
% on an expected utility outcomes as a strategy for action choice in uncertain
% environments \citep{gershman2015computational}.
In our present model-free RL framework,
a rat might represent such knowledge about the world as follows:
\begin{itemize}
\item $r(s, \text{``stand still''}) = 0$ if $s$ does not correspond to a cell / chamber containing food.
\item $p(s'|s,\text{``stand still''}) = 1$ if $s'=s$ and $0$ otherwise.
\item etc.
\end{itemize}

@@

\subsection{The Q-values, state-values, and the advantage function}
We can now formulate the purpose of the DMN
as finding a policy $\pi$ for the human agent that maximizes the
expected cumulative value of an state-action pair $(a,s)$, also known as the \textit{Q-value}, namely
\begin{equation}
  \label{eq:qval}
  Q^{\pi}(s,a) = \mathbb E [G^\pi|s_0=s,a_0=a].
\end{equation}
In other words,
the Q-value $Q^{\pi}(s,a)$ corresponds to the expected reward
over all possible action trajectories, in which
the agent sets out in the environment in state
$s$, chooses action $a$,
and then follows the policy $\pi$ to select future actions.
%% Given that each $\pi$ correspondence to an exhaustive description of
%% behavioral patterns, $Q^{\pi}(s_t,a_t)$ corresponds to the epistemic value of
%% the set of actions given states in finite time.
For the brain,
$Q^{\pi}(s, a)$ defined in \eqref{eq:qval} provides the subjective
utility of executing a specific action; it answers the question
"What is the expected utility of taking action $a$ in this situation ?".
$Q^{\pi}(s,a)$ offers a formalization of optimal behavior that
may well reflect processing aspects of the DMN in human agents.

One can also define the state-value function

\begin{eqnarray}
  \V^\pi(s) := \mathbb E_{a \sim \pi}Q^\pi(s,a) = \mathbb E [G^\pi|s_0=s],
\end{eqnarray}
which measures the value of each state $s \in \mathcal S$ under the policy $\pi$. This serves
as a baseline value for any action. Finally, the advantage function is defined by
\begin{equation}
A^\pi(s,a) :=  Q^\pi(s,a) - V^\pi(s).
\end{equation}

@@

\subsection{Efficient computation of optimal behavior}
By construction, maximizing the expectation in  \eqref{eq:qval} is computationally intractable in general.
This is due to the enormous number of possible action
trajectories that the agent often needs to
consider simultaneously.
% \begin{mdframed}
%   \textbf{N.B.:}
%   If the Bellman-based computational ideas have a corresponding
%   neurobiological implementation, this provides a framework of how
%   optimal behavior could be computable in the human brain.
% \end{mdframed}
One popular solution is called \text{Q-learning}. It remedies
the issue by only considering the sub-family of
\textit{deterministic policies}, which map each state to the single best action to
take at the current state. Such policies define deterministic functions from states to actions.

\subsubsection{Q-learning} Q-learning (Watkins \& Dayan, 1992) generates a
\textit{greedy policy}
\textbf{XXX please add 1-2 sentences why this is interesting in the current
context. This statement appears a little alone as it stands.}
\begin{equation}
  \pi^*(s) = \argmax_{a \in \mathcal A}{Q}^*(s, a),
  \label{eq:qlearning}
\end{equation}
which maps each state $s$ to the optimal action which should be taken at that state.
Here $Q^*(s, a) := \max_{\pi}Q^\pi(s,a)$, the optimal Q-value over all policies.
% where $\tilde{Q}(s, a)$ is the current estimate of the state-action value
% table.
The \textit{Bellman equation} \citep{sutton1998reinforcement} then takes the simple form
\begin{equation}
  \begin{split}
    Q^*(s_0, a_0) &= \mathbb E_{s_1 \sim p(s_1|s_0,a)} [r(s,a) + \gamma \max_{a_1 \in \mathcal A}Q^*(s_1, a_1)]
    \\
    &= \mathbb E_{s_1 \sim p(s_1|s_0,a_0)} [r(s_0,a_0) + \gamma Q^*(s_1, \pi(s_1))].
  \end{split}
  \label{eq:bellman}
\end{equation}
This consistency equation for the optimal (i.e., over all policies) Q-value function provides
a recursive decomposition of optimal behavior by dividing the Q-value function into the immediate
reward and the discounted rewards of the upcoming states. The optimal Q-value operator $Q^*$
is a fixed point for this equation. A complicated dynamic programming optimization can be
broken into simpler sub-problems at different time points.
Exploitation of hierarchical structure in actions considerations
has previously been closely related to the medial prefrontal part of the DMN
\citep{koechlin1999role, braver2002role}.
Using the Bellman equation, each state can be associated with a certain value
to guide action towards a better state, thus improving on the agent's current policy.
% It is a fixed-point equation for the deterministic policy $\pi$.
Note that in \eqref{eq:bellman} the sampling is done only over things which
depend on the environment, and so can be learned off-policy by observing state transitions
triggered by another behavioral policy, which can be stochastic.

\subsubsection{Efficient learning via value approximation and experience replay}
%As already mentioned in the previous section, Q-learning optimizes over the class of
% deterministic policies of the form \eqref{eq:qlearning}.
A practically successful RL approach
needs ways to efficiently approximate the state-action value function
to allow for scaling to large state and action spaces.
In a state $s \in \mathcal S$, the agent takes an action $a \in \mathcal A$
% (for example, $a$ = ``Read'', or ``Run'', or ``Laugh'', or ``Sympathize'', or ``Empathize'', etc.)
by sampling from its current policy, and collects a reward $r$,
% (measured on their internal currency of worldly values ?),
after which the environment transitions to a new state $s' \in \mathcal S$.
% \textbf{XXX Please try to smoothen the previous sentence}
At the end of this cycle, a new \textit{experience} $e = (s,a,r,s')$ is produced, which represents an exemplar
behavior of the agent and is recorded in replay memory buffer
(possibly discarding the oldest entries to make space):
$\mathcal D \leftarrow \text{append}(\mathcal D, e)$.
At iteration $k+1$, replay then consists in sampling (uniform or importance-weighted
\footnote{e.g weighted by TD error of the state transition $s \overset{a}{\rightarrow} s'$.} ?)  mini-batches of experiences
$(s, a, r, s') \sim \mathcal D$ from the replay memory $\mathcal D$.
The agent then tries to approximate
the would-be $Q$-value for the state-action pair $(s,a)$ as predicted by the Bellman equation \eqref{eq:bellman}, namely
\begin{equation}
  y_k := y_k(s,a,s') =  r + \gamma \max_{a'} \tilde{Q}(s', a'|\theta_{k-1}),
\end{equation}
with the prediction of a parametrized regression model $(s,a)
\mapsto \tilde{Q}(s, a|\theta_{k-1})$. Here the function $(s,a) \mapsto \tilde{Q}(a,s|\theta_{k-1})$
% \textbf{XXX just double-checking, are the dots int he Q() correct?}
is an approximator for the $Q$-value operator,  parameterized by weights $\theta_{k}$
(synaptic strengths, etc.).

From a neurobiological perspective,
experience replay would be manifested as the re-occurrence of a
neuron spiking sequence that also occurs during specific actions.
%, but the replay has a much faster time scale.
In the human brain, the hippocampus is likely to contribute to such a mechanism,
as neuroscience experiments on various animals have repeatedly indicated
in rats, mice, cats, rabbits, songbirds, and
monkeys \cite{buhry2011,nokia2010,dave2000,skaggs2007}.

Computing an optimal policy then corresponds to finding the parameters $\theta_{k}$ which minimize the following mean-squared loss function
\begin{equation}
  \mathcal L(\theta^Q_{k})
  % = \mathbb E_{(s, a, r, s') \sim \mathcal D}[(\tilde{Q}(s, a|\theta) - \tilde{Q}(s',a'|\theta^Q_{k}))^2]
  = \mathbb E_{(s, a, r, s') \sim \mathcal D}\left[\frac{1}{2}(\tilde{Q}(s, a|\theta_{k}) - y_k)^2\right],
  \label{eq:oracle}
\end{equation}
where $y_k$ is a defined in \eqref{eq:bellman}
% \textbf{We still may need to explain y in the formula with one sentence.}
For instance, a general linear model with a kernel $\phi$ would be of the
form
$$\tilde{Q}(s, a|\theta) = \phi(s,a)^T\theta.$$
$\phi(s,a)$ would represent a high-level hand-crafted representation of the state-action pairs
$(s,a)$, as proposed by \citep{songNIPS2016} for example.
%  and Parr et al. 2003 (JMLR)].
A recently proposed, practically successful alternative approach
\citep{mnih2015,silver2016mastering} is to learn this
representation using a deep neural network, leading to the so-called \textit{Deep Q-learning} family of methods which
likely are state-of-the-art in RL.
The set of $\theta$ parameters that instantiate the nonlinear interactions
between layers of the artifical neural network
may find a neurobiological correspondence in the adaptive strengths of axonal
connections between neurons from the different levels
of the neural processing hierarchy
\citep{mesulam1998sensation, taylor2015global}.

Learning
of the the entire model parameters can effectively be achieved
via \textit{backpropagation},
where reward prediction errors (RPEs) -- also known as \textit{regret}-- percolate
from lower to higher processing layers to modulate the choice of future actions
% online stochastic gradient descent is used to minimize the variational objective
% \eqref{eq:oracle}.
